Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-05-11 latex/vanishing.tex
Some minor changes, plus an explanation of how to modify clustered to non clustered

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}T
The necessary conditions for D-optimality of Theorem \ref{thm:char}
leave some freedom in choosing $\obs$, which may or may not give rise
clustered designs. Consider an inverse problem for which the first two
eigenvalues of $\fwd \prcov\fwd^*$ are $\lambda_1^{-1} = 0.1,
\lambda_2^{-1} = 0.2$, its eigenvectors are $\{\ev_i\}_{i=1}^\infty$
and $\sigma^2=1$. Consider the following $\obs_1, \obs_2$ (see also
Figure \ref{fig:clusterization}):
\begin{equation}\label{eq:two designs}
  \obs_1 =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \obs_2 =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}
\noindent It is easy to verify that $\obs_1^*\obs_1 = \obs_2^*\obs_2$
and that both $\obs_1$ and $\obs_2$ are D-optimal. However, $\obs_1$
is a clustered design ($\meas_3 = \meas_5 = \ev_1$), while $\obs_2$ is
not.

@@

\textbf{Question \ref{q:why}}: Since $\obs_1$ is clustered and
D-optimal, measurement clusterization poses no obstruction to
D-optimality. Furthermore, a computer implementation of Lemma
\ref{thm:char} also generates $\obs_1$ as a solution to the D-optimal
design problem. Randomized numerical simulations give rise to
D-optimal designs that exhibit clustrization more than 95\% of the
time (see code in supplementary). Since our model is as generic as
possible, we conclude that it is possible that measurement
clusterization is a generic phenomenon that.

\textbf{Question \ref{q:replace}}: Since $\obs_2$ is D-optimal but not
clustered, we conclude that it is possible to substitute a clustered
design for an equally optimal non-clustered design. Achieveing this in
the setup of our model not hard, but in real life problems this may
prove more difficult.

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs_1$ in \eqref{eq:two designs})
    and non-clusterization (right, $\obs_2$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs_1$ and $\obs_2$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because neither $\meas_1$ nor
    $\meas_2$ are not in the direction of any eigenvector. Similarly
    for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}

>>>>

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
The necessary conditions for D-optimality of Theorem \ref{thm:char}
leave some freedom in choosing $\obs$, which may or may not give rise
clustered designs. Consider an inverse problem for which the first two
eigenvalues of the pushforward prior $\fwd \prcov\fwd^*$ are
$\lambda_1 = 10, \lambda_2 = 5$, its eigenvectors are
$\{\ev_i\}_{i=1}^\infty$ and $\sigma^2=1$. Consider the following
$\obs, \pobs$ (see also Figure \ref{fig:clusterization}):
\begin{equation}\label{eq:two designs}
  \obs =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \pobs =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}
\noindent It is easy to verify that $\obs^*\obs = \pobs^*\pobs$
and that both $\obs$ and $\pobs$ are D-optimal. However, $\obs$
is a clustered design ($\meas_3 = \meas_5 = \ev_1$), while $\pobs$ is
not.

@@

\textbf{Question \ref{q:why}}: Since $\obs$ is clustered and
D-optimal, measurement clusterization poses no obstruction to
D-optimality. Furthermore, a computer implementation of Lemma
\ref{thm:char} also generates $\obs$ as a solution to the D-optimal
design problem. Randomized numerical simulations give rise to
D-optimal designs that exhibit clustrization more than 95\% of the
time (see code in supplementary). Since our model is as generic as
possible, we conclude that it is possible that measurement
clusterization is a generic phenomenon that.

\textbf{Question \ref{q:replace}}: Since $\pobs$ is D-optimal but not
clustered, we conclude that it is possible to substitute a clustered
design for an equally optimal non-clustered design. It is easy to give
some simple intuition on how the clustered design $\obs$ was modified
to the non-clustered design $\pobs$. First let us introduce some
notation: let $\obs_{12}$ be the design composed of the first and
second rows of $\obs$ and use a similar definition for
$\pobs_{12}$. Note that $\obs_{12}$ uniquely determines $\obs$ as a
clustered design (up to permutations and signs). Thus, we wanted to
slightly modify $\obs_{12}$. We replaced $\sqrt{11/40}$, the
coefficent of $\ev_1$ in the first measurement of $\obs$, with
$\sqrt{3/8}$ in the first measurement of $\pobs$. After that, we chose
$\pobs_{12}$ such that its measuremtns have unit norm and
$\pobs_{12}^* \pobs_{12}$ has the same eigenvectors as
$\obs^*\obs$. We then considered a ``new'' inverse problem, for which
the pushforward prior covariance is $(\fwd \prcov\fwd^*)^{-1} +
\sigma^{-2} \pobs_{12}^*\pobs_{12}$. We then completed $\pobs$ to a
D-optimal design via the construction detailed in Lemma
\ref{lemma:free}. This is how $\pobs$ was constructed, but
(infinitely) many other non-clustered D-optimal designs could be
constructed this way and it is not clear why the numerical
implementation of Lemma \ref{lemma:free} resulted in a clustered
design. Giving a full algorithm for the procedure detailed above is
out of scope for the current study, and I intend to pursue it in the
future.

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs$ in \eqref{eq:two designs})
    and non-clusterization (right, $\pobs$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs$ and $\pobs$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because neither $\meas_1$ nor
    $\meas_2$ are not in the direction of any eigenvector. Similarly
    for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}
